Memory for television episodes preserves event content while introducing new across-event similarities

### 2019-12-11 paper/main.tex
updated brain methods

\subsection*{Modeling the dynamic content of the video and recall transcripts}
\subsubsection*{Topic modeling}
The input to the topic model we trained to characterize the dynamic content of the video comprised 998 hand-generated annotations of short (mean: 2.96s) scenes spanning the video clip~(\citealp{ChenEtal17} generated 1000 annotations total; we removed two referring to the break between the first and second scan sessions, during which no fMRI data was collected).  The features annotated included: narrative details (a sentence or two describing what happened in that scene); whether the scene took place indoors or outdoors; names of any characters that appeared in the scene; name(s) of characters in camera focus; name(s) of characters who were speaking in the scene; the location (in the story) that the scene took place; camera angle (close up, medium, long, top, tracking, over the shoulder, etc.); whether music was playing in the scene or not; and a transcription of any on-screen text.  We concatenated the text for all of these features within each segment, creating a ``bag of words'' describing each scene.  We then re-organized the text descriptions into overlapping sliding windows of 50 scenes each.  In other words, we created a ``context" for each scene comprising the text descriptions of the preceding 25 scenes, the present scene, and the following 24 scenes.  To model the ``context" at the beginning and end of the video (i.e., within 50 scenes of the beginning or end), we created overlapping sliding windows that grew in size from one scene to the full length, then similarly tapered their length at the end.  This bore the additional benefit of representing each scene's description in the text corpus an equal number of times.

@@

We used an optimization procedure to select the appropriate $K$ for each topic proportions matrix.  Specifically, we computed (for each matrix)
\[
  \argmax_K \left[POM(\frac{a}{b-J}) - \frac{K}{\alpha}\right],
\]
where $a$ was the average correlation between the topic vectors of timepoints within the same state; $b$ was the average correlation between the topic vectors of timepoints within \textit{different} states; \textit{J} was a constant used to ensure a positive denominator, set equal to $\min_K\left[{\frac{a}{b}}\right]$; and $\alpha$ was a regularization parameter that we set to 5 times the window length (i.e., 250 scenes for the video topic trajectory and 50 sentences for the recall topic trajectories).  Before subtracting the regularization term, we scaled the ratio of correlations to be a proportion of the maximum ratio across all $K$'s.  Figure~\ref{fig:model}B displays the event boundaries returned for the video, and Figure~\corrmats~displays the event boundaries returned for each participant's recalls (See Fig.~\kcorrs~for the values of \textit{a} and \textit{b} for each $K$, Fig.~\kopt~for the optimization functions for the video and recalls).  After obtaining these event boundaries, we created stable estimates of each topic proportions matrix by averaging the topic vectors within each event.  This yielded a number-of-events by number-of-topics matrix for the video and recalls from each participant.

We also evaluated a parameter-free procedure for choosing $K$, which finds the $K$ value that maximizes the Wasserstein distance (a.k.a. ``Earth mover's" distance) between the within and across event distributions of correlation values. This alternative procedure largely replicated the pattern of results found with the parameterized method described above, but recovered substantially fewer events on average (Fig.\wasserstein). While both approaches seem to underestimate the number of video/recall events relative to the ``true" number (as determined by human raters), the parameterized approach was closer to the true number.

@@

Perhaps the simplest and most widely used measure of memory performance is \textit{accuracy}---i.e., the proportion of studied (experienced) items (in this case, the 34 video events) that the participant later remembered.  \cite{ChenEtal17} developed a human rating system whereby the quality of each participant's memory was evaluated by an independent rater.  We found a strong across-participants correlation between these independant ratings and the overall number of events that our HMM approach identified in participants' recalls (Pearson's $r(15) = 0.64, p = 0.006$).

@@

\paragraph{Probability of first recall (PFR).}  PFR curves~\citep{WelcBurn24, PostPhil65, AtkiShif68} reflect the probability that an item will be recalled first as a function of its serial position during encoding. To carry out this analysis, we initialized a number-of-participants (17) by number-of-video-events (34) matrix of zeros. Then for each participant, we found the index of the video event that was recalled first (i.e., the video event whose topic vector was most strongly correlated with that of the first recall event) and filled in that index in the matrix with a 1.  Finally, we averaged over the rows of the matrix, resulting in a 1 by 34 array representing the proportion of participants that recalled an event first, as a function of the order of the event's appearance in the video (Fig.~\ref{fig:list-learning}A).

\paragraph{Lag conditional probability curve (lag-CRP).} The lag-CRP curve~\citep{Kaha96} reflects the probability of recalling a given event after the just-recalled event, as a function of their relative positions (or \textit{lag}).  In other words, a lag of 1 indicates that a recalled event came immediately after the previously recalled event in the video, and a lag of -3 indicates that a recalled event came 3 events before the previously recalled event.  For each recall transition (following the first recall), we computed the lag between the current recall event and the next recall event, normalizing by the total number of possible transitions.  This yielded a number-of-participants (17) by number-of-lags (-33 to +33; 67 lags total) matrix. We averaged over the rows of this matrix to obtain a group-averaged lag-CRP curve (Fig.~\ref{fig:list-learning}B).

\paragraph{Serial position curve (SPC).} SPCs~\citep{Murd62a} reflect the proportion of participants that remember each item as a function of the items' serial position during encoding. We initialized a number-of-participants (17) by number-of-video-events (34) matrix of zeros. Then, for each recalled event, for each participant, we found the index of the video event that the recalled event most closely matched (via the correlation between the events' topic vectors) and entered a 1 into that position in the matrix (i.e., for the given participant and event). This resulted in a matrix whose entries indicated whether or not each event was recalled by each participant (depending on whether the corresponding entires were set to one or zero).  Finally, we averaged over the rows of the matrix to yield a 1 by 34 array representing the proportion of participants that recalled each event as a function of the order of the event's appearance in the video (Fig.~\ref{fig:list-learning}C).

@@

% NOTE: Thinking about removing this analysis? More of a sanity check than something novel and cool, and takes up space. Happy to be convinced otherwise
% leaving this one to Jeremy's judgement
% -Paxton

@@

\subsubsection*{Visualizing the video and recall topic trajectories}
We used the UMAP algorithm~\citep{McInHeal18} to project the 100-dimensional topic space onto a two-dimensional space for visualization (Figs.~\ref{fig:trajectory}, \ref{fig:topics}).  To ensure that all of the trajectories were projected onto the \textit{same} lower dimensional space, we computed the low-dimensional embedding on a ``stacked'' matrix created by vertically concatenating the events-by-topics topic proportions matrices for the video and all 17 participants' recalls.  We then divided the rows of the result (a total-number-of-events by two matrix) back into separate matrices for the video topic trajectory and the trajectories for each participant's recalls (Fig.~\ref{fig:trajectory}).  This general approach for discovering a shared low-dimensional embedding for a collections of high-dimensional observations follows \cite{HeusEtal18a}.

@@

\subsection*{Searchlight fMRI analyses}
In Figure~\ref{fig:brainz}, we present two analyses aimed at identifying brain structures whose responses (as participants viewed the video) exhibited particular temporal correlations.  We developed a searchlight analysis whereby we constructed a cube centered on each voxel (radius: 5~voxels).  For each of these cubes, we computed the temporal correlation matrix of the voxel responses during video viewing.  Specifically, for each of the 1976 volumes collected during video viewing, we correlated the activity patterns in the given cube with the activity patterns (in the same cube) collected during every other timepoint.  This yielded a 1976 by 1976 correlation matrix for each cube.

Next, we constructed two sets of ``template'' matrices: one reflecting the video's topic trajectory and the other reflecting each participant's recall topic trajectory.  To construct the video template, we computed the correlations between the topic proportions estimated for every pair of TRs (prior to segmenting the trajectory into discrete events; i.e., the correlation matrix shown in Figs.~\ref{fig:model}B and \ref{fig:brainz}A).  We constructed similar temporal correlation matrices for each participant's recall topic trajectory (Figs.~\ref{fig:model}D, \corrmats).  However, to correct for length differences and potential non-linear transformations between viewing time and recall time, we first used dynamic time warping~\citep{BernClif94} to temporally align participants' recall topic trajectories with the video topic trajectory (an example correlation matrix before and after warping is shown in Fig.~\ref{fig:brainz}B).  This yielded a 1976 by 1976 correlation matrix for the video template and for each participant's recall template.

To determine which (cubes of) voxel responses reliably matched the video template, we correlated the upper triangle of the voxel correlation matrix for each cube with the upper triangle of the video template matrix~\citep{KrieEtal08b}.  This yielded, for each participant, a single correlation value.  We computed the average (Fisher $z$-transformed) correlation coefficient across participants.  We used a permutation-based procedure to assess significance, whereby we re-computed the average correlations for each of 100 ``null'' video templates (constructed by circularly shifting the template by a random number of timepoints).  (For each permutation, the same shift was used for all participants.)  We then estimated a $p$-value by computing the proportion of shifted correlations that were larger than the observed (unshifted) correlation.  To create the map in Figure~\ref{fig:brainz}A we thresholded out any voxels whose correlation values fell below the 95\textsuperscript{th} percentile of the permutation-derived null distribution.

We used a similar procedure to identify which voxels' responses reflected the recall templates.  For each participant, we correlated the upper triangle of the correlation matrix for each cube of voxels with their (time warped) recall correlation matrix.  As in the video template analysis this yielded a single correlation coefficient for each participant.  However, whereas the video analysis compared every participant's responses to the same template, here the recall templates were unique for each participant.  We computed the average $z$-transformed correlation coefficient across participants, and used the same permutation procedure we developed for the video responses to assess significant correlations.  To create the map in Figure~\ref{fig:brainz}B we thresholded out any voxels whose correlation values fell below the 95\textsuperscript{th} percentile of the permutation-derived null distribution.

>>>>

\subsection*{Modeling the dynamic content of the video and recall transcripts}
\subsubsection*{Topic modeling}
The input to the topic model we trained to characterize the dynamic content of the video comprised 998 hand-generated annotations of short (mean: 2.96s) scenes spanning the video clip~(\citealp{ChenEtal17} generated 1000 annotations total; we removed two referring to the break between the first and second scan sessions, during which no fMRI data was collected).  The features annotated included: narrative details (a sentence or two describing what happened in that scene); whether the scene took place indoors or outdoors; names of any characters that appeared in the scene; name(s) of characters in camera focus; name(s) of characters who were speaking in the scene; the location (in the story) that the scene took place; camera angle (close up, medium, long, top, tracking, over the shoulder, etc.); whether music was playing in the scene or not; and a transcription of any on-screen text.  We concatenated the text for all of these features within each segment, creating a ``bag of words'' describing each scene.  We then re-organized the text descriptions into overlapping sliding windows of 50 scenes each.  In other words, we created a ``context" for each scene comprising the text descriptions of the preceding 25 scenes, the present scene, and the following 24 scenes.  To model the ``context" at the beginning and end of the video (i.e., within 25 scenes of the beginning or end), we created overlapping sliding windows that grew in size from one scene to the full length, then similarly tapered their length at the end.  This bore the additional benefit of representing each scene's description in the text corpus an equal number of times.
% ------ NOTE: ------
% need to refine explanation here a little bit
% - Paxton

@@

We used an optimization procedure to select the appropriate $K$ for each topic proportions matrix.  Prior studies on narrative structure and processing have shown that we both perceive and internally represent the world around us at multiple, hierarchical timescales (e.g., \citealp{HassEtal08, LernEtal11, HassEtal15, ChenEtal17, BaldEtal17, BaldEtal18}).  However, for the purposes of our framework, we sought to identify the single timescale of event-representations that is emphasized \textit{most heavily} in the temporal structure of the video and each participant's recalls.  We quantified this as the set of $K$ event boundaries that yielded the maximal distinctiveness between the content (i.e., topics) within each event and that in all other events. Specifically, we computed (for each matrix)
\[
  \argmax_K \left[W_{1}(a, b)\right],
\]
where $a$ was the distribution of correlations between the topic vectors of timepoints within the same state and $b$ was the average correlation between the topic vectors of timepoints within \textit{different} states.  For each possible $K$, we computed the first Wasserstein distance ($W_{1}$; also known as ``earth mover's distance"; \citealp{Dobr70, RamdEtal17}) between these distributions, and chose the $K$-value that yielded the greatest difference.  Figure~\ref{fig:model}B displays the event boundaries returned for the video, and Figure~\corrmats~displays the event boundaries returned for each participant's recalls (See Fig.~\kopt~for the optimization functions for the video and recalls).  After obtaining these event boundaries, we created stable estimates of each topic proportions matrix by averaging the topic vectors within each event.  This yielded a number-of-events by number-of-topics matrix for the video and recalls from each participant.

@@

Perhaps the simplest and most widely used measure of memory performance is \textit{accuracy}---i.e., the proportion of studied (experienced) items (in this case, the 30 video events) that the participant later remembered.  \cite{ChenEtal17} developed a human rating system whereby the quality of each participant's memory was evaluated by an independent rater.  We found a strong across-participants correlation between these independant ratings and the overall number of events that our HMM approach identified in participants' recalls (Pearson's $r(15) = 0.65, p = 0.004$).

@@

\paragraph{Probability of first recall (PFR).}  PFR curves~\citep{WelcBurn24, PostPhil65, AtkiShif68} reflect the probability that an item will be recalled first as a function of its serial position during encoding. To carry out this analysis, we initialized a number-of-participants (17) by number-of-video-events (30) matrix of zeros. Then for each participant, we found the index of the video event that was recalled first (i.e., the video event whose topic vector was most strongly correlated with that of the first recall event) and filled in that index in the matrix with a 1.  Finally, we averaged over the rows of the matrix, resulting in a 1 by 30 array representing the proportion of participants that recalled an event first, as a function of the order of the event's appearance in the video (Fig.~\ref{fig:list-learning}A).

\paragraph{Lag conditional probability curve (lag-CRP).} The lag-CRP curve~\citep{Kaha96} reflects the probability of recalling a given event after the just-recalled event, as a function of their relative positions (or \textit{lag}).  In other words, a lag of 1 indicates that a recalled event came immediately after the previously recalled event in the video, and a lag of -3 indicates that a recalled event came 3 events before the previously recalled event.  For each recall transition (following the first recall), we computed the lag between the current recall event and the next recall event, normalizing by the total number of possible transitions.  This yielded a number-of-participants (17) by number-of-lags (-29 to +29; 61 lags total) matrix. We averaged over the rows of this matrix to obtain a group-averaged lag-CRP curve (Fig.~\ref{fig:list-learning}B).

\paragraph{Serial position curve (SPC).} SPCs~\citep{Murd62a} reflect the proportion of participants that remember each item as a function of the items' serial position during encoding. We initialized a number-of-participants (17) by number-of-video-events (30) matrix of zeros. Then, for each recalled event, for each participant, we found the index of the video event that the recalled event most closely matched (via the correlation between the events' topic vectors) and entered a 1 into that position in the matrix (i.e., for the given participant and event). This resulted in a matrix whose entries indicated whether or not each event was recalled by each participant (depending on whether the corresponding entires were set to one or zero).  Finally, we averaged over the rows of the matrix to yield a 1 by 30 array representing the proportion of participants that recalled each event as a function of the order of the event's appearance in the video (Fig.~\ref{fig:list-learning}C).

@@

% NOTE: Thinking about removing this analysis? More of a sanity check than something novel and cool, and takes up space. Happy to be convinced otherwise - andy
% leaving this one to Jeremy's judgement
% -Paxton

@@

\subsubsection*{Visualizing the video and recall topic trajectories}
We used the UMAP algorithm~\citep{McInHeal18} to project the 100-dimensional topic space onto a two-dimensional space for visualization (Figs.~\ref{fig:trajectory}, \ref{fig:topics}).  To ensure that all of the trajectories were projected onto the \textit{same} lower dimensional space, we computed the low-dimensional embedding on a ``stacked'' matrix created by vertically concatenating the events-by-topics topic proportions matrices for the video and all 17 participants' recalls.  We then divided the rows of the result (a total-number-of-events by two matrix) back into separate matrices for the video topic trajectory and the trajectories for each participant's recalls (Fig.~\ref{fig:trajectory}).  This general approach for discovering a shared low-dimensional embedding for a collections of high-dimensional observations follows \cite{HeusEtal18a}.
% ------ NOTE: ------
% need to add info about embedding optimization
% - Paxton

@@

\subsection*{Searchlight fMRI analyses}
In Figure~\ref{fig:brainz}, we present two analyses aimed at identifying brain regions whose responses (as participants viewed the video) exhibited a particular temporal structure.  We developed a searchlight analysis wherein we constructed a cube centered on each voxel (radius: 5~voxels) and for each of these cubes, computed the temporal correlation matrix of the voxel responses during video viewing.  Specifically, for each of the 1976 volumes collected during video viewing, we correlated the activity patterns in the given cube with the activity patterns (in the same cube) collected during every other timepoint.  This yielded a 1976 by 1976 correlation matrix for each cube.

Next, we constructed a series of ``template'' matrices: the first reflecting the timecourse of video's topic trajectory, and the others reflecting that of each participant's recall topic trajectory.  To construct the video template, we computed the correlations between the topic proportions estimated for every pair of TRs (prior to segmenting the trajectory into discrete events; i.e., the correlation matrix shown in Figs.~\ref{fig:model}B and \ref{fig:brainz}A).  We constructed similar temporal correlation matrices for each participant's recall topic trajectory (Figs.~\ref{fig:model}D, \corrmats).  However, to correct for length differences and potential non-linear transformations between viewing time and recall time, we first used dynamic time warping~\citep{BernClif94} to temporally align participants' recall topic trajectories with the video topic trajectory.  An example correlation matrix before and after warping is shown in Fig.~\ref{fig:brainz}B.  This yielded a 1976 by 1976 correlation matrix for the video template and for each participant's recall template.

To determine which (cubes of) voxel responses matched the video template, we correlated the upper triangle of the voxel correlation matrix for each cube with the upper triangle of the video template matrix~\citep{KrieEtal08b}.  This yielded, for each participant, a voxelwise map of correlation values.  We then performed a one-sample $t$-test on the distribution of (Fisher $z$-transformed) correlations at each voxel, across participants.  This resulted in a value for each voxel (cube), describing how reliably its timecourse mirrored that of the video.

We further sought to ensure that our analysis identified regions where the activations’ temporal structure specifically reflected that of the video, rather than regions whose activity was simply autocorrelated at a width similar to the video template’s diagonal.  To achieve this, we used a phase shift-based permutation procedure, wherein we circularly shifted the video’s topic trajectory by a random number of timepoints, computed the resulting ``null” video template, and re-ran the searchlight analysis, in full.  (For each of the 100 permutations, the same random shift was used for all participants).  We $z$-scored the observed (unshifted) result at each voxel against the distribution of permutation-derived ``null” results, and estimated a $p$-value by computing the proportion of shifted results that yielded larger values.  To create the map in Figure~\ref{fig:brainz}A, we thresholded out any voxels whose similarity to the unshifted video’s structure fell below the 95\textsuperscript{th} percentile of the permutation-derived similarity results.

We used an analogous procedure to identify which voxels' responses reflected the recall templates.  For each participant, we correlated the upper triangle of the correlation matrix for each cube of voxels with their (time warped) recall correlation matrix.  As in the video template analysis this yielded a voxelwise map of correlation coefficients per participant.  However, whereas the video analysis compared every participant's responses to the same template, here the recall templates were unique for each participant.  As in the analysis described above, we $t$-scored the (Fisher $z$-transformed) voxelwise correlations, and used the same permutation procedure we developed for the video responses to ensure specificity to the recall timeseries and assign significance values.  To create the map in Figure~\ref{fig:brainz}B we again thresholded out any voxels whose correspondence values fell below the 95\textsuperscript{th} percentile of the permutation-derived null distribution.
